```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

---
title: "Analyzing the effect of transmition type on MPG efficiency"
author: "Jamamel"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MotorTrendMPG}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Objective

The following analysis identifies alternative scenarios in MPG efficiency in cars due to the presence of different types of transmition (automatic vs. manual). These are described through statistical inference and provide a quantifiable measure of benefit from one type compared to the other. All analyses use R dataset *mtcars*, and combine a series of exploratory and regression based techniques to quantify efficiency based on observed data.

We use dataset *mtcars* from the *datasets* R package, and our aim is to determine the impact of several variables in the efficiency and accuracy of quantifying fuel consumption efficiency through mile-per-galon (*MPG*). Specifically, two questions will be answered:

1. Is an automatic or manual transmission better for *MPG*?
2. How much better (measured by change in *MPG*) is one type of transmition vs. the other? 

We start by identifying possible correlation among variables included in our data. A series of exploratory analyses helped identify data relationship candidates. Specifically, any relationships to transmission type (automatic vs. manual; variable *am*) & mpg (variable *mpg*). 

```{r exploratory, echo=FALSE, warning=FALSE, results='hide', cache=TRUE}
library(ggplot2)
library(GGally)
library(dplyr)
data(mtcars)

mtcars$am <- factor(mtcars$am, levels = c(0,1), labels = c('automatic', 'manual'))
mtcars$vs <- factor(mtcars$vs, levels = c(0,1), labels = c('v', 'straight'))
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
# mtcars$cyl <- factor(mtcars$cyl)
mtcars$cwt <- scale(mtcars$wt, scale = FALSE)
mtcars$cdisp <- scale(mtcars$disp, scale = FALSE)
mtcars$toy <- ifelse(rownames(mtcars) == 'Toyota Corolla',1,0)
mtcars$fiat <- ifelse(rownames(mtcars) == 'Fiat 128',1,0)


contcorr <- ggpairs(mtcars,
        colour = 'am',
        columns = c(1,3:7,9))

disccorr <- ggpairs(mtcars,
        colour = 'am',
        columns = c(1:2,8:11))
```

*mtcars* contains data for `r nrow(mtcars)` unique car models, with the following measures:

Measure (Variable) | Definition
- | -
mpg | Miles per gallon
cyl | Number of cylinders
disp | Displacement (cubic inches)
hp | Gross horsepower
drat | Rear axle ration
wt | Weight (lb/1000)
qsec | Quarter mile time
vs | V vs. Straight engine
am | Transmission (automatic vs. manual)
gear | Number of forward gears
carb | Number of carburetors

# Exploratory Analysis

A combination of correlation, and plotting is performed for initial exploration.

*mpg* has high correlation with a few variables relating to the car dimensions. Also, a car's weight (*wt*), displacement (*disp*), and gross horse power (*hp*) are themselves highly correlated. This suggests any linear-based regression modelling approach will confound any effects. This is especially true should they all be included in a model directly (see *Figure 1*). There also appear to be clear relationships between transmision type (*am*) and attributes like number of cylinders (*cyl*) or carburetors (*carb*), or engine type (*vs*). However, the distribution of the number of forward gears (*gear*) is less clear, perhaps due to the small number of observations in our dataset (see *Figure 2*).

While some patterns are visible between transmission and some other attributes, it is sometimes difficult to determine whether these patterns are logical and expected, rather than purely random observations in this particular data. So, while manual transmission cars tend to have more forward gears, cylinders, and carburetors, there's no immediate evidence of these being linked more directly to engine type.

# Modelling MPG

Next, a few of the relationships identified were modelled through linear regression to try and quantify the effect of different car specs on *MPG*, namely transmission type. The exploratory phase confirmed linear regression was a suitable approach given *mpg* empirically close to a Normal distribution.

After a series of iterations (described in comments of [R Markdown file](https://github.com/Jamamel/MotorTrendMPG/blob/master/R/MotorTrendMPG.R)), the best model was that in which *mpg* was described by the interaction between type of transmision and weight, number of cylinders (as a factor referencing 4 cylinders, the type of transmision on its own, and two dummy variables for Toyota Corolla and Fiat 128 cars. These 2 cars were modelled uniquely given their leverage on the model (see results below). 

```{r models, echo=FALSE, results='hide', fig.keep='none', warning=FALSE}
# First model reveals a possible relationship of wt and hp when all other effects present. 
# Good fit but mostly driven by large number of parameters. 
fit0 <- lm(mpg ~ ., mtcars)
summary(fit0)

# Large residuals and the assumption of Normality in the regressions error terms appears not to be accurate.
par(mfrow = c(2,2))
plot(fit0)

# By defining a model where MPG is explained by factor effect of transmition type and weight along with their interaction. Number of cylinders is also introduced as factor.
fit <- lm(mpg ~ am * cwt + cyl + toy + fiat, mtcars)
summary(fit)

# Adjustment is not much worse than when all variables (and its high number of parameters) were used. In addition, all coefficients have low enough standard errors to be acceptable for the author (especially given the low sample size given the complex relationships present and the limitation of tools allowed in this exercise). 



# Lower fit, greater residual variance, less nice QQ plot. General lower accuracy observed.
fit2 <- lm(mpg ~ am + cyl, mtcars)
summary(fit2)
par(mfrow = c(2,2))
plot(fit2)


# Still a mediocre model with slightly lower residual variance than fit2. The rogue group of cars is larger, meaning we don't describe enough patterns with as much consistency.
fit3 <- lm(mpg ~ am + cyl + wt, mtcars)
summary(fit3)
par(mfrow = c(2,2))
plot(fit3)

```

```{r fitmodel, echo=FALSE}
knitr::kable(round(summary(fit)$coefficients, 3))
```

All estimated coefficients have acceptable p-values, and adjusted overall fit stands at a high **`r round(summary(fit)$adj.r.squared, 2)` with `r summary(fit)$df[2]` degrees of freedom** . Of more interest is the fact the residual standard error is low at **`r round(summary(fit)$sigma,2)`**. Residual analysis also shows all residuals are within 1.5 standard deviations of the mean. Normality is fairly consistent apart from a couple of cases which don't have enough leverage on the analysis to make a meaningful difference (*See Figure 3*).

# Results & Conclusions
 
At average car weight, manual transmision on its own appears to have a slight negative effect on MPG (`r round(summary(fit)$coefficients['ammanual',1], 2)` vs. the equivalent automatic transmision). This effect, however, is mostly compensated by other usual attributes automatic transmision cars which greatly hinder MPG. For every ton above the average, MPG decreases by `r round(summary(fit)$coefficients['cwt',1], 2)`, and for every extra cylinder, MPG decreases by a further `r round(summary(fit)$coefficients['cwt',1], 2)`. When one considers the average difference in weight and usually higher number of cylinders, the counter effect becomes clear. 

\newpage

# Appendix

```{r contcorr, anchor=TRUE, echo=FALSE, warning=FALSE, results='markup', message=FALSE, fig.cap='Correlation of car dimensions to mpg split by transmision type.'}
print(contcorr)
```

```{r disccorr, anchor=TRUE, echo=FALSE, warning=FALSE, results='markup', message=FALSE, fig.cap='Distribution of car attributes to and mpg split by transmision type.'}
print(disccorr)
```

```{r plotmodels, anchor=TRUE, echo=FALSE, warning=FALSE, results='markup', message=FALSE, fig.cap='Model Fit Diagnostics'}
# fit0 - Large residuals and the assumption of Normality in the regressions error terms appears not to be accurate.
par(mfrow = c(2,2))
plot(fit0)


# fit - Residual variance is lower and the assumption of Normality, while not perfect, is better founded. Four cars have a similar patterns that aren't easily isolated with simple linear regression.  However, fit and parsimony of the model were deemed acceptable.
par(mfrow = c(2,2))
plot(fit)
```
