---
title: "Analyzing the effect of transmition type on MPG efficiency"
author: "Jamamel"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MotorTrendMPG}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Objective

The following analysis identifies alternative scenarios in MPG efficiency in cars due to the presence of different types of transmition (automatic vs. manual). These are described through statistical inference and provide a quantifiable measure of benefit from one type compared to the other. All analyses use R dataset *mtcars*, and combine a series of exploratory and regression based techniques to quantify efficiency based on observed data.

We use dataset *mtcars* from the *datasets* R package, and our aim is to determine the impact of several variables in the efficiency and accuracy of quantifying fuel consumption efficiency through mile-per-galon (*MPG*). Specifically, two questions will be answered:

1. Is an automatic or manual transmission better for *MPG*?
2. How much better (measured by change in *MPG*) is one type of transmition vs. the other? 

We start by identifying possible correlation among variables included in our data. A series of exploratory analyses helped identify data relationship candidates. Specifically, any relationships to transmission type (automatic vs. manual; variable *am*) & mpg (variable *mpg*). 

```{r, echo=FALSE, warning=FALSE, results='hide'}
library(ggplot2)
library(GGally)
data(mtcars)

mtcars$am <- factor(mtcars$am, levels = c(0,1), labels = c('automatic', 'manual'))
mtcars$vs <- factor(mtcars$vs, levels = c(0,1), labels = c('v', 'straight'))
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
mtcars$cyl <- factor(mtcars$cyl)

contcorr <- ggpairs(mtcars,
        colour = 'am',
        columns = c(1,3:7,9))

disccorr <- ggpairs(mtcars,
        colour = 'am',
        columns = c(1:2,8:11))
```

*mtcars* contains data for `r nrow(mtcars)` unique car models, with the following measures:

Measure (Variable) | Definition
- | -
**mpg | Miles per gallon**
cyl | Number of cylinders
disp | Displacement (cubic inches)
hp | Gross horsepower
drat | Rear axle ration
wt | Weight (lb/1000)
qsec | Quarter mile time
vs | V vs. Straight engine
**am | Transmission (automatic vs. manual)**
gear | Number of forward gears
carb | Number of carburetors

# Exploratory Analysis

A combination of correlation, and plotting is performed for initial exploration.

*mpg* has high correlation with a few variables relating to the car dimensions. Also, a car's weight (*wt*), displacement (*disp*), and gross horse power (*hp*) are themselves highly correlated. This suggests any linear-based regression modelling approach will confound any effects. This is especially true should they all be included in a model directly (see *Figure 1*). There also appear to be clear relationships between transmision type (*am*) and attributes like number of cylinders (*cyl*) or carburetors (*carb*), or engine type (*vs*). However, the distribution of the number of forward gears (*gear*) is less clear, perhaps due to the small number of observations in our dataset (see *Figure 2*).

While some patterns are visible between transmission and some other attributes, it is sometimes difficult to determine whether these patterns are logical and expected, rather than purely random observations in this particular data. So, while manual transmission cars tend to have more forward gears, cylinders, and carburetors, there's no immediate evidence of these being linked more directly to engine type.

# Modelling MPG

Next, a few of the relationships identified were modelled through linear regression to try and quantify the effect on *MPG* of different car specs, namely transmission type. The exploratory phase confirmed linear regression was a suitable approach given *mpg* empirically close to a Normal distribution.



\newpage

# Appendix

```{r contcorr, anchor=TRUE, echo=FALSE, warning=FALSE, results='markup', message=FALSE, fig.cap='Correlation of car dimensions to mpg split by transmision type.'}
print(contcorr)
```

```{r disccorr, anchor=TRUE, echo=FALSE, warning=FALSE, results='markup', message=FALSE, fig.cap='Distribution of car attributes to and mpg split by transmision type.'}
print(disccorr)
```
